Synthesizing Symmetric Lenses

### 2018-05-29 papers/SymmetricLenses/main.tex
add some more impl etc.

\section{Synthesis Algorithm}

% begin stoch-rx
\section{Stochastic Regular Expressions}
\label{sec:stoch-rx}
Important to our algorithm is finding the expected number of bits needed to
recover a string in one data source from a synchronized string in the other data
source. To do this, we must first develop a probability model for our language.
We do this with \emph{stochastic regular expressions}, regular expressions
annotated with probability information. With this, we can jointly express a
language and a probability distribution over that language.

@@

\begin{theorem}
  If $\Regex \SSREquiv \RegexAlt$ then $\ProbabilityOf{\Regex}{\String} =
  \ProbabilityOf{\RegexAlt}{\String}$, for all strings $\String \in \LanguageOf{\Regex}$.
\end{theorem}
% end stoch-rx

% begin DNF regular expressions and lenses
\section{Stochastic DNF Regular Expressions}
\label{sec:dnf-regex-lens}
We extend stochastic regular expressions to stochastic DNF regular expressions.
\begin{center}
  \begin{tabular}{l@{\ }c@{\ }l@{\ }>{\itshape\/}r}
    % DNF_REGEX
    \Atom{},\AtomAlt{} & \GEq{} & \PRegexStar{\DNFRegex{}}{\Probability}
% & \StarAtomType{}
\\
    \Sequence{},\SequenceAlt{} & \GEq{} &
                                                       $\SequenceOf{\String_0\SeqSep\Atom_1\SeqSep\ldots\SeqSep\Atom_n\SeqSep\String_n}$ 
%& \MultiConcatSequenceType{} 
\\
    \DNFRegex{},\DNFRegexAlt{} & \GEq{} & $\DNFOf{(\Sequence_1,\Probability_1)\DNFSep\ldots\DNFSep(\Sequence_n,\Probability_n)}$ %& \MultiOrDNFRegexType{} 
  \end{tabular}
\end{center}

@@

This means we also have an algorithm for finding the entropy of an unambiguous
stochastic regular expression, as we can merely convert the regular expression
into DNF form, and find the entropy of it.
% end DNF regular expressions and lenses

% begin symmetric dnf lenses
\section{Symmetric DNF Lenses}
\label{sec:symmetric-dnf-lenses}
\subsection{Syntax}
Symmetric DNF Lenses are what we synthesize, and we then turn them into
Symmetric Lenses. The nonterminals $i$, $j$, $p$, $q$, $r$, $c$, and $d$ all
represent natural numbers. The nonterminals $\String$ and $\StringAlt$ represent
strings.

\begin{center}
  \begin{tabular}{@{}r@{\ }c@{}l@{}}
    % REGEX
    \SAtomLens{} & \GEq{} & $\IterateLensOf{\SDNFLens}$ \\
    \SSQLens{} & \GEq{} & $(\SSQLensOf{(i_1,j_1,\SAtomLens_1)\SeqLSep
                          \ldots\SeqLSep
                          (i_p,j_p,\SAtomLens_p)}
                          ,\ListOf{(\String_1,\Atom_1);\ldots;(\String_q,\Atom_q)}
                          ,\ListOf{\StringAlt_1;\ldots;\StringAlt_r})$ \\
    \SDNFLens{} & \GEq{} & $(\SDNFLensOf{(i_1,j_1,\SSQLens_1)\DNFLSep
                           \ldots\DNFLSep
                           (i_p,j_p,\SSQLens_p)}
                           ,\ListOf{c_1;\ldots;c_q}
                           ,\ListOf{d_1;\ldots;d_r})$ \\
  \end{tabular}
\end{center}

@@

\subsection{Typing and Semantics}
Because these lenses only encode functions on well-typed programs, we introduce
the semantics alongside the typing rules.  The semantics are over typing
derivations, instead of over individual lenses.

\subsection{DNF Lenses}
The typing judgment is a 3-ary relation over a single DNF lens, and two DNF
regular expressions. If $\SDNFLens \OfType \DNFRegex \Leftrightarrow
\DNFRegexAlt$, then the $\SDNFLens.\CreateR$, $\SDNFLens.\CreateL$,
$\SDNFLens.\PutR$, and $\SDNFLens.\PutL$ functions form a symmetric lens.

@@

\paragraph*{Sequence Lenses}
The typing judgment is a 3-ary relation over a single Sequence lens, and two
sequences. If $\SSQLens \OfType \Sequence \Leftrightarrow \SequenceAlt$, then
the $\SSQLens.\CreateR$, $\SSQLens.\CreateL$, $\SSQLens.\PutR$, and
$\SSQLens.\PutL$ functions form a symmetric lens.

@@

\paragraph*{Atom Lenses}
The typing judgment is a 3-ary relation over a single atom lens, and two
atoms. If $\SAtomLens \OfType \Atom \Leftrightarrow \AtomAlt$, then
the $\SAtomLens.\CreateR$, $\SAtomLens.\CreateL$, $\SAtomLens.\PutR$, and
$\SAtomLens.\PutL$ functions form a symmetric lens.

@@

\[
  \inferrule*
  {
    \SDNFLens \OfType \DNFRegex \Leftrightarrow \DNFRegexAlt
  }
  {
    \IterateLensOf{\SDNFLens}
    \OfType \PRegexStar{\DNFRegex}{\Probability}
    \Leftrightarrow \PRegexStar{\DNFRegexAlt}{\ProbabilityAlt}
  }
\]

@@

Now that we've developed symmetric DNF lenses, it's important to develop a means
to guage how bijective the lens is.

@@

% end symmetric dnf lenses

% begin synthesis
\section{Synthesis}
\label{sec:synthesis}
While examples can be used to tighten the
specification, a large number of well-constructed examples are required to
sufficiently constrain the search space.

Unfortunately, trying to find the best symmetric lens is a daunting task!
Unless we find a bijective lens (in which case the sum is zero), we can never be
sure if we will find a better lens than the one we found.
% end synthesis

% begin implementation
\section{Implementation}
\label{sec:implementation}
% end implementation

@@

% begin related-work
\section{Related Work}
\label{sec:related}

@@

These components only make sense for synchronization when they satisfy
round-tripping laws.  This guarantees that if an identity edit is propagated, no
data gets changed.
\begin{equation}
  \tag{\PutRL}
  \frac{\PutRSymOf{(x,c)} = (y,c')}{\PutLSymOf{(y,c'')} = (x,c')}
\end{equation}
\begin{equation}
  \tag{\PutLR}
  \frac{\PutLSymOf{(y,c)} = (x,c')}{\PutRSymOf{(x,c')} = (y,c')}
\end{equation}

>>>>

% begin synthesis
\section{Synthesis}
In this work, we aim to quickly find a good lens between two regular expressions
that satisfies the input-output examples. To this end, we develop an information
theoretic metric for how much data is synchronized. Information theoretic
metrics require a notion of probability distributions \emph{how are probability
  distributions being used, are certain things weighted more etc, assign weights
  to different subcomponents and parts of the data}. We annotate
the regular expressions with a reasonable probability distribution, making them
\emph{stochastic regular expressions}. Armed with probability distributions over
the languages, we have the capabilities to define information theoretic metrics
over these probability distributions.

Our overall synthesis algorithm utilizes two synthesizers shown in
Figure~\ref{fig:high-level-algorithm}. There are an infinite number of possible
lenses between two regular expressions that can be searched through. Instead of
searching through all possible lenses, the lenses are grouped into classes of
lenses based on the syntax of the regular expressions they synchronize. Two
lenses are in the same class if they can be typed by the same regular
expressions, using no star equivalences. This works well for synthesis, as
finding a good lens within these classes permits an efficient synthesis
procedure.

However, proposing these classes is complicated by stochastic regular
expressions. While the equivalences on regular expressions are well studied, the
equivalences on stochastic regular expressions are not. We want to propose
equivalent stochastic regular expressions, but both the languages and
probability distributions must be maintained. If the probability distributions
are not maintained, then the metric by which we compare our lenses changes for
each class of lenses, we can't compare lenses between classes. To alleviate
this, we describe the \emph{star-semiring} equivalences on stochastic regular
expressions, and prove they are sound with respect to language and probability
distributions. Our \RXSearch proposes stochastic regular expressions by
traversing a subset of these equivalences, so \RXSearch only proposes regular
expression pairs equivalent to the originals.

The second algorithm, \GreedySynth, searches within a class for the lowest cost
lens.  It is a greedy algorithm, and not guaranteed to find the lowest cost
lens.  However, we haven't found any instances in which \GreedySynth doesn't find
the optimal lens, due to the partial backtracking we implemented. \GreedySynth
works by searching, not through symmetric lenses, but through an alternative
language of Symmetric DNF lenses. After finding such a symmetric DNF lens, that
symmetric DNF lens is converted into a simple symmetric lens.

\subsection{Stochastic Regular Expressions}
\label{subsec:stoch-rx}
Important to our algorithm is finding the expected number of bits needed to
recover a string in one data source from a synchronized string in the other data
source. To do this, we must first develop a probability model for our language.
We do this with \emph{stochastic regular expressions}, regular expressions
annotated with probability information. With this, we can jointly express a
language and a probability distribution over that language.

@@

\begin{theorem}
  If $\Regex \SSREquiv \RegexAlt$ then $\ProbabilityOf{\Regex}{\String} =
  \ProbabilityOf{\RegexAlt}{\String}$, for all strings $\String \in \LanguageOf{\Regex}$.
\end{theorem}

\subsection{\RXSearch}
The first synthesizer is one that searches through regular expression pairs.
This synthesizer, \RXSearch, proposes regular expressions, semantically but not
syntactically equivalent to the originals.  In particular, this synthesizer
performs the following rewrites:
\begin{center}
  \begin{tabular}{rcl}
    $\PRegexStar{\Regex}{\Probability}$
    & \Rewrite
    & \PRegexOr{\EmptyString{}}{(\RegexConcat{\Regex{}}{\PRegexStar{\Regex{}}{\Probability}})}{\Probability}\\

    $\PRegexStar{\Regex}{\Probability}$
    & \Rewrite
    & \PRegexOr{\EmptyString{}}{(\RegexConcat{\PRegexStar{\Regex{}}{\Probability}}{\Regex{}})}{\Probability}
  \end{tabular}
\end{center}
As well as the congruence rules that allow these rewrites to be applied on
subexpressions.

\RXSearch proposes rewrites incrementally, where it proposes regular expression
pairs with the smallest number of rewrites first, and when there are no more
proposed regular expressions of that size, \RXSearch proposes regular
expressions of larger size.

\RXSearch terminates when the number of regular expressions of the smallest
unproposed size exceeds the cost of the current best lens. At this point,
\RXSearch stops proposing regular expression pairs, and instead returns to the
user the best lens found thus far.  Note, if \RXSearch finds a bijective lens (a
lens of cost 0), it will immediately return, as would be expected.

\afm{dont forget to add in description of termination}
%

\subsection{\GreedySynth}
The second synthesizer, \GreedySynth, actually generates the lenses. It does
this by first converting the regular expressions into \emph{stochastic DNF
  regular expressions}, intuitively stochastic regular expressions with all
disjunctions distributed over all concatenations. Next, a good DNF lens is
greedily found between these stochastic DNF regular expressions, using the
probability annotations on the DNF regular expressions to find the cost.
Finally, this DNF lens is converted into a simple symmetric lens.

\paragraph*{Stochastic DNF Regular Expressions}
\label{sec:dnf-regex-lens}
We extend stochastic regular expressions to stochastic DNF regular expressions.
\begin{center}
  \begin{tabular}{l@{\ }c@{\ }l@{\ }>{\itshape\/}r}
    % DNF_REGEX
    \Atom{},\AtomAlt{} & \GEq{} & \PRegexStar{\DNFRegex{}}{\Probability}
% & \StarAtomType{}
\\
    \Sequence{},\SequenceAlt{} & \GEq{} &
                                                       $\SequenceOf{\String_0\SeqSep\Atom_1\SeqSep\ldots\SeqSep\Atom_n\SeqSep\String_n}$ 
%& \MultiConcatSequenceType{} 
\\
    \DNFRegex{},\DNFRegexAlt{} & \GEq{} & $\DNFOf{(\Sequence_1,\Probability_1)\DNFSep\ldots\DNFSep(\Sequence_n,\Probability_n)}$ %& \MultiOrDNFRegexType{} 
  \end{tabular}
\end{center}

@@

This means we also have an algorithm for finding the entropy of an unambiguous
stochastic regular expression, as we can merely convert the regular expression
into DNF form, and find the entropy of it.

\paragraph*{Symmetric DNF, Sequence, and Atom Lenses}
\label{sec:symmetric-dnf-lenses}
Symmetric DNF Lenses are what we synthesize, and we then turn them into
Symmetric Lenses. The nonterminals $i$, $j$, $p$, $q$, $r$, $c$, and $d$ all
represent natural numbers. The nonterminals $\String$ and $\StringAlt$ represent
strings.

\begin{center}
  \begin{tabular}{@{}r@{\ }c@{}l@{}}
    % REGEX
    \SAtomLens{} & \GEq{} & $\StarOf{\SDNFLens}$ \\
    \SSQLens{} & \GEq{} & $(\SSQLensOf{(i_1,j_1,\SAtomLens_1)\SeqLSep
                          \ldots\SeqLSep
                          (i_p,j_p,\SAtomLens_p)}
                          ,\ListOf{(\String_1,\Atom_1);\ldots;(\String_q,\Atom_q)}
                          ,\ListOf{\StringAlt_1;\ldots;\StringAlt_r})$ \\
    \SDNFLens{} & \GEq{} & $(\SDNFLensOf{(i_1,j_1,\SSQLens_1)\DNFLSep
                           \ldots\DNFLSep
                           (i_p,j_p,\SSQLens_p)}
                           ,\ListOf{c_1;\ldots;c_q}
                           ,\ListOf{d_1;\ldots;d_r})$ \\
  \end{tabular}
\end{center}

@@

Because these lenses only encode functions on well-typed programs, we introduce
the semantics alongside the typing rules.  The semantics are over typing
derivations, instead of over individual lenses.

\paragraph*{Symmetric DNF Lenses, Typing and Semantics}
The typing judgment is a 3-ary relation over a single DNF lens, and two DNF
regular expressions. If $\SDNFLens \OfType \DNFRegex \Leftrightarrow
\DNFRegexAlt$, then the $\SDNFLens.\CreateR$, $\SDNFLens.\CreateL$,
$\SDNFLens.\PutR$, and $\SDNFLens.\PutL$ functions form a symmetric lens.

@@

\paragraph*{Sequence Lenses, Typing and Semantics}
The typing judgment is a 3-ary relation over a single Sequence lens, and two
sequences. If $\SSQLens \OfType \Sequence \Leftrightarrow \SequenceAlt$, then
the $\SSQLens.\CreateR$, $\SSQLens.\CreateL$, $\SSQLens.\PutR$, and
$\SSQLens.\PutL$ functions form a symmetric lens.

@@

\paragraph*{Atom Lenses, Typing and Semantics}
The typing judgment is a 3-ary relation over a single atom lens, and two
atoms. If $\SAtomLens \OfType \Atom \Leftrightarrow \AtomAlt$, then
the $\SAtomLens.\CreateR$, $\SAtomLens.\CreateL$, $\SAtomLens.\PutR$, and
$\SAtomLens.\PutL$ functions form a symmetric lens.

@@

\[
  \inferrule*
  {
    \SDNFLens \OfType \DNFRegex \Leftrightarrow \DNFRegexAlt
  }
  {
    \StarOf{\SDNFLens}
    \OfType \PRegexStar{\DNFRegex}{\Probability}
    \Leftrightarrow \PRegexStar{\DNFRegexAlt}{\ProbabilityAlt}
  }
\]

@@

\paragraph*{Information-Theoretic Cost Metric}
Now that we've developed symmetric DNF lenses, it's important to develop a means
to guage how bijective the lens is.

@@

\paragraph*{Greedy Algorithm}
The greedy algorithm is actually two greedy algorithms, one that greedily finds
symmetric DNF lenses and one that greedily finds symmetric sequence lenses.

The algorithm for symmetric DNF lenses is relatively straightforward.  First,
using examples, we identify which atoms in one format \emph{must} be
synchronized to certain atoms in the other.  After that, we iteratively find the
pairs of atoms with lowest cost, where at least one of the atoms is not mapped
to another atom.  The cost of these pairs is not merely the cost of the lens,
but also the cost incurred by adding that lens, if one of the atoms is already
involved in a map, the lens has additional cost to account for the merge cost.
This continues until there are no more atoms that aren't mapped to another atom.
At this point, the creates for each lens is determined.  If a examples dictate
one lens must create to another, that is the create behavior, otherwise atoms
create along the lens of least cost.

The algorithm for symmetric sequence lenses is less straightforward.  Unlike
symmetric DNF lenses, where multiple subcomponents can map to the same
subcomponent, each atom can map to at most one other subcomponent.  This is
complicated by the fact that examples sometimes dictate one atom must map to
another.

To address this, we introduce the \emph{profit} of a adding a mapping between
two subcomponents. The profit is the cost of the lens, minus the information
content of the subcomponents it maps between, plus how much profit would be lost
by freeing those subcomponents to make the subcomponents available. Then, lenses
are only added if they have positive profit, or if one of the atoms requires a
mapping. After there are no more lenses with positive profit or with
subomponents requiring a mapping, the sequence lens is returned.

\subsection{Optimizations}

We included a number of optimizations in our algorithm. We added ``open'' and
``closed'' annotations to regular expressions, an expansion inference algorithm,
hash-consing and memoization, and compositional synthesis to our algorithm.

\paragraph*{Open and Closed Regular Expressions} While GreedySynth acts
relatively efficiently, it suffers from an exponential blowup when converting
regular expressions to DNF form. This can be minimized by not converting
\emph{all} of the regular expression into DNF form, but only converting
\emph{some} of it to DNF form. We add in an additional atom, where an atom can
be an either an iteration of a DNF regular expression or it can be a regular
expression. When converted into a DNF regular expression, if $\Regex$ is a
stochastic regular expression, annotated as closed, then $\ToDNFRegexOf{\Regex}
= \AtomToDNFOf{\Regex}$.

At the start of synthesis, we annotate our regular expressions as closed. Now,
we enhance our rewrites by changing an annotation to from closed to open. This
forces changes to \RXSearch and \GreedySynth.  \RXSearch is changed in that now
it can perform new rewrites.  \GreedySynth is changed in that now there is an
additional type of atom that can be mapped between.  Two regular expressions can
map to each other only through the identity lens: two regular expressions can
map to each other by the identity lens, if they are the same and all the
examples say they are the same, otherwise each regular expression is left
disconnected.

\paragraph*{Expansion Inference} The prior optimization helped \GreedySynth, but
it made \RXSearch have a much harder job.  To alleviate this difficulty, we
added in expansion inference.  Expansion inference identifies when a regular
expression does not have a chance to map through any means other than
disconnected.  Expansion inference opens those regular expressions.

\paragraph*{Hash-consing and Memoization} We further help \GreedySynth by
memoizing its results.  However, our stochastic DNF regular expressions are
quite large and so hashing them is computationally intensive.  We use less
memory and allow for quick comparisons and hashing of these DNF regular
expressions by hash-consing them.

\paragraph*{Compositional Synthesis} Compositional synthesis allows us to use
previously defined lenses - whether they are previously defined by users or by
synthesis. We alter \GreedySynth such that two regular expression atoms can be
mapped to each other either if they are the same regular expression and have
examples agreeing with the examples, or if there is a lens which is a
composition of predefined lenses between them, where the lens alters the
examples in a desired way.
% end synthesis

% begin implementation
\section{Implementation}
We have implemented our symmetric DNF lenses and the synthesis of them as an
extension to Boomerang. We added symmetric lens combinators to Boomerang, and
implemented Boomerang's previous asymmetric lens combinators within the
symmetric lens combinators.

We have extended Optician with our symmetric synthesis algorithm.
% end implementation

@@

% begin related-work
\section{Related Work}
\label{sec:related}
There is related work in a number of directions.  There is related work in
synthesizing synchronization functions, related work on symmetric lenses,
related work on string transformation synthesis, and related work on
type-directed synthesis.

\subsection{Symmetric Lenses}
We have reformulated the problem addressed by symmetric lenses with simple
symmetric lenses.

@@

These components only make sense for synchronization when they satisfy
round-tripping laws.  This guarantees that if an identity edit is propagated, no
data gets changed.
\begin{equation}
  \tag{\PutRL}
  \frac{\PutRSymOf{(x,c)} = (y,c')}{\PutLSymOf{(y,c''''')} = (x,c')}
\end{equation}
\begin{equation}
  \tag{\PutLR}
  \frac{\PutLSymOf{(y,c)} = (x,c''')}{\PutRSymOf{(x,c')} = (y,c')}
\end{equation}
